Supporting Information

### 2018-03-30 SupportingInformation.Rmd
figure presntn changed

csl: science.csl
bibliography: REF_SI.bib
---

@@

The third step of model selection was to fit multiple bivariate copula families to the ranked data $(u_i,v_i)$ via maximum likelihood, using the `BiCopEst` function in `VineCopula`. Families considered (with abbreviations and numeric codes here indicated) included the **N**ormal/Gaussian {1}, **C**layton {3}, **G**umbel {4}, **F**rank {5}, **J**oe {6}, **BB1** {7}, **BB6** {8}, **BB7** {9}, **BB8** {10}, **S**urvival **C**layton {13}, **S**urvival **G**umbel {14}, **S**urvival **J**oe {16}, **S**urvival **BB1** {17}, **S**urvival **BB6** {18}, **S**urvival **BB7** {19} and **S**urvival **BB8** {20} copulas.
Akaike Information Criterion (AIC) and Bayesian Information Criterion (BIC) values and accompanying model weights ($\text{AIC}_{\text{w}}$, $\text{BIC}_{\text{w}}$) were generated using standard formulas @burnham2003_modelselection. 
Descriptions of these copula families can be found in introductory texts on copulas @nelsen2007_copula , @joe2014_dependence and in references cited in `VineCopula`. The lower- and upper-tail dependence of the 
best-fitting member of each family was also provided by `BiCopEst`.

@@

Finally, $\text{AIC}_{\text{w}}$ values for all fitted models were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas @burnham2003_modelselection; likewise for BIC.

@@

Left- and right-tail dependence quantify the extent to which two ranked
variables $u_i$, $v_i$ for $i=1,...,n$ 
(see Sec. \ref{Model_selection} for defintions of $u_i$ and $v_i$)
are related in the left and right tails, respectively, of their 
distributions @nelsen2007_copula. 
Our nonparametric study of tail dependence uses three
statistics which can quantify the extent to which $u_i$ and $v_i$ are 
related in any part of their distributions. We here describe the 
statistics. The statistics were formulated with positively associated
variables in mind. All the variables we study are positively associated
when they are significantly associated. So definitions and associated 
diagrams will assume positive associations, though generalizations are 
possible. Given two bounds 
$0 \leq l_b < u_b \leq 1$, we define the parallel lines $u+v=2l_b$ and 
$u+v=2u_b$ which intersect the unit square in which the $u_i$ and $v_i$ 
fall (these are the blue and red lines, respectively, in 
Figure \ref{fig_stat_image}). Our statistics quantify the dependence
between $u_i$ and $v_i$ in the region bounded by these lines.
Using $l_b=0$ and some value $u_b\leq 0.5$ can give information about 
dependence in the left parts of the distributions of $u$ and $v$, and 
using $u_b=1$ and $l_b \geq 0.5$ can give information about dependence in 
the right parts of the distributions.

@@

The difference $\cor_l-\cor_u$ is expected to be positive if 
dependence in the left halves of the distributions is stronger 
than dependence in their right halves; likewise for $\Ps_l-\Ps_u$.
This holds, for instance, for the Clayton copula @nelsen2007_copula.
The difference $\Dsq_u-\Dsq_l$, in which the "upper" and
"lower" statistics occur in opposite order, is also expected to be 
positive under the same conditions. We use these three differences
to detect asymmetry of tail dependence. Likewise, for large datasets,
we consider $\cor_{l_b,u_b}-\cor_{1-u_b,1-l_b}$, 
$\Ps_{l_b,u_b}-\Ps_{1-u_b,1-l_b}$ and
$\Dsq_{1-u_b,1-l_b}-\Dsq_{l_b,u_b}$ with $l_b=0$ and $u_b$ close to $0$.

In addition to [basic unit tests](https://github.com/sghosh89/BIVAN/blob/master/Tests.Rmd), we 
also tested our statistics by testing whether $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and 
$\Dsq_u-\Dsq_u$ revealed the asymmetry of tail dependence known to be present in data simulated
from various copulas. Normal and Frank copula families 
are known to have symmetric tail dependence,
i.e., strengths of dependence in the lower and upper tails are the same, for all copulas
in these families @nelsen2007_copula. Clayton copulas are known to have stronger lower- than 
upper-tail dependence. And survival Clayton copulas, being rotations of Clayton copulas by 
$180^\circ$, have stronger upper- than lower tail dependence @nelsen2007_copula.
Using the `iRho` function from the `copula` package, for each of the four families normal, 
Frank, Clayton, and survival Clayton, we obtained parameters for which the expected value of 
Spearman's $\rho$ was, separately, $0,0.1,0.2,...,0.9$. For each copula family and for each 
of the selected parameter values we generated  $n$ points, 100 times, and thereby computed 100 values of the 
statistics $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_u$. This exercise 
was repeated, separately, 
for $n=1000$ and for $n=35$, these cases being chosen, respectively, to reflect asymptotic behavior of the 
statistics and their behavior on the small datasets which are common in some fields
of ecology. The mean and standard error of the three difference statistics were computed for 
each copula family, parameter value, and value of $n$, and results were plotted against 
Spearman's rho (Figures \ref{fig_stat_testing35} and \ref{fig_stat_testing1000}), 
the expectation being that the difference statistics would be positive for 
Clayton copulas and negative for survival Clayton copulas (except for the case of $\rho=0$, a 
boundary case of these families corresponding to independence), and not significantly different 
from $0$ (we used a $t$-test with a significance threshold of $0.05$) for Frank and normal 
copulas. These expectations were fullfilled (Figures \ref{fig_stat_testing35} and 
\ref{fig_stat_testing1000}).

@@

For multivariate datasets, e.g., the aphid abundance or first flight data, each of which consists of time series 
from 10 locations, a spatial resampling scheme was used instead of the surrogate approach described above.
The scheme is identical to that proposed by @bjornstadfalck2001 for application in their now 
widely used spline correlogram method; the same scheme was also used by @walteretal2017 as part of a method for 
performing model selection among matrix regression models. For each pair of locations $i \neq j$, the statistic
of interest, here denoted $s_{ij}$ ($\cor_{l_b,u_b}$, $\Ps_{l_b,u_b}$, $\Dsq_{l_b,u_b}$, $\cor_l-\cor_u$, 
$\Ps_l-\Ps_u$, or $\Dsq_u-\Dsq_l$) was computed for those locations. The values $s_{i,j}$ were then arranged in
a matrix, $S$, with diagonal entries NA. The mean of the non-NA entries of $S$ was computed. For 
$\cor_{l_b,u_b}$, $\Ps_{l_b,u_b}$, or $\Dsq_{l_b,u_b}$ this mean characterized the average strength of 
dependence, across all pairs of distinct locations, for data in the portion of the distributions 
given by $l_b$ and $u_b$. For $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, or $\Dsq_u-\Dsq_l$, the mean characterized average
asymmetry of tail dependence across all pairs of locations. 
Confidence intervals for the values of $\mean(S)$ were then computed as quantiles of a distribution of bootstrapped values.
Bootstrapping was performed by resampling the locations of measurement, with replacement; this corresponds to 
resampling, with replacement, the rows and columns of $S$, and is described in detail elsewhere 
@bjornstadfalck2001, @walteretal2017. 
<!--Shya, these citations render as (4), (5), whereas they should render as (4, 5), or better still, (Bjornstad & Falck, 2001; Walter et al, 2017) (the author-date callout style is better when space is not a limitation because it gives the reader more information without having to look in the bibliography). Can you fix this, please?-->

@@

<!--Shya, I dont think we need this, so I have commented it out. If you disagree, let me know, otherwise
please go ahead and delete it and this comment.
![copula plot for aphid-count\label{fig_count_allcop}](./Results/stat_results/stat_aphid_count/Sp_10_AllCops.pdf)
-->

<!--Shya, can you please organize these following plots into one plot with mutliple panels? Once that is done, please add some text describing these results.-->
![Plot for aphid-count : Spearman vs. Distance\label{fig_count_spear}](./Results/stat_results/stat_aphid_count/Sp_10_Spearman_vs_D.pdf){ width=50%, height=50% }

![Plot for aphid-count : Kendall vs. Distance\label{fig_count_kend}](./Results/stat_results/stat_aphid_count/Sp_10_Kendall_vs_D.pdf){ width=50%, height=50% }

![$cor_l$-$cor_u$ vs. Distance plot for aphid-count showing lower tail dependence\label{fig_count_Corl-Coru}](./Results/stat_results/stat_aphid_count/Sp_10_Corl-Coru_vs_D.pdf){ width=50%, height=50% }

![$P_l$-$P_u$ vs. Distance plot for aphid-count showing lower tail dependence\label{fig_count_Pl-Pu}](./Results/stat_results/stat_aphid_count/Sp_10_Pl-Pu_vs_D.pdf){ width=50%, height=50% }

![$D^2_u$-$D^2_l$ vs. Distance plot for aphid-count showing lower tail dependence\label{fig_count_D2u-D2l}](./Results/stat_results/stat_aphid_count/Sp_10_D2u-D2l_vs_D.pdf){ width=50%, height=50% }

@@

<!--As for the count data, I don't think we need this. If you disagree, let me know, otherwise pls delete the commented text and this comment.
![copula plot for aphid-firstflight\label{fig_ff_allcop}](./Results/stat_results/stat_aphid_ff/Sp_11_AllCops.pdf)
-->

<!--Shya, can you please organize these following plots into one plot with mutliple panels? Once that is done, please add some text describing these results.-->
![Plot for aphid-firstflight : Spearman vs. Distance\label{fig_ff_spear}](./Results/stat_results/stat_aphid_ff/Sp_11_Spearman_vs_D.pdf){ width=50%, height=50% }

![Plot for aphid-firstflight : Kendall vs. Distance\label{fig_ff_kend}](./Results/stat_results/stat_aphid_ff/Sp_11_Kendall_vs_D.pdf){ width=50%, height=50% }

![$cor_l$-$cor_u$ vs. Distance plot for aphid-firstflight showing upper tail dependence\label{fig_ff_Corl-Coru}](./Results/stat_results/stat_aphid_ff/Sp_11_Corl-Coru_vs_D.pdf){ width=50%, height=50% }

![$P_l$-$P_u$ vs. Distance plot for aphid-firstflight showing upper tail dependence\label{fig_ff_Pl-Pu}](./Results/stat_results/stat_aphid_ff/Sp_11_Pl-Pu_vs_D.pdf){ width=50%, height=50% }

![$D^2_u$-$D^2_l$ vs. Distance plot for aphid-firstflight showing upper tail dependence\label{fig_ff_D2u-D2l}](./Results/stat_results/stat_aphid_ff/Sp_11_D2u-D2l_vs_D.pdf){ width=50%, height=50% }

## Results: abundance data for *Ceratium furca* in UK seas \label{Ex_pns}

>>>>

#csl: science.csl
bibliography: REF_SI.bib
---

@@

The third step of model selection was to fit multiple bivariate copula families to the ranked data $(u_i,v_i)$ via maximum likelihood, using the `BiCopEst` function in `VineCopula`. Families considered (with abbreviations and numeric codes here indicated) included the **N**ormal/Gaussian {1}, **C**layton {3}, **G**umbel {4}, **F**rank {5}, **J**oe {6}, **BB1** {7}, **BB6** {8}, **BB7** {9}, **BB8** {10}, **S**urvival **C**layton {13}, **S**urvival **G**umbel {14}, **S**urvival **J**oe {16}, **S**urvival **BB1** {17}, **S**urvival **BB6** {18}, **S**urvival **BB7** {19} and **S**urvival **BB8** {20} copulas.
Akaike Information Criterion (AIC) and Bayesian Information Criterion (BIC) values and accompanying model weights ($\text{AIC}_{\text{w}}$, $\text{BIC}_{\text{w}}$) were generated using standard formulas [@burnham2003_modelselection]. 
Descriptions of these copula families can be found in introductory texts on copulas [@nelsen2007_copula ; @joe2014_dependence]  and in references cited in `VineCopula`. The lower- and upper-tail dependence of the 
best-fitting member of each family was also provided by `BiCopEst`.

@@

Finally, $\text{AIC}_{\text{w}}$ values for all fitted models were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.

@@

Left- and right-tail dependence quantify the extent to which two ranked
variables $u_i$, $v_i$ for $i=1,...,n$ 
(see Sec. \ref{Model_selection} for defintions of $u_i$ and $v_i$)
are related in the left and right tails, respectively, of their 
distributions [@nelsen2007_copula]. 
Our nonparametric study of tail dependence uses three
statistics which can quantify the extent to which $u_i$ and $v_i$ are 
related in any part of their distributions. We here describe the 
statistics. The statistics were formulated with positively associated
variables in mind. All the variables we study are positively associated
when they are significantly associated. So definitions and associated 
diagrams will assume positive associations, though generalizations are 
possible. Given two bounds 
$0 \leq l_b < u_b \leq 1$, we define the parallel lines $u+v=2l_b$ and 
$u+v=2u_b$ which intersect the unit square in which the $u_i$ and $v_i$ 
fall (these are the blue and red lines, respectively, in 
Figure \ref{fig_stat_image}). Our statistics quantify the dependence
between $u_i$ and $v_i$ in the region bounded by these lines.
Using $l_b=0$ and some value $u_b\leq 0.5$ can give information about 
dependence in the left parts of the distributions of $u$ and $v$, and 
using $u_b=1$ and $l_b \geq 0.5$ can give information about dependence in 
the right parts of the distributions.

@@

The difference $\cor_l-\cor_u$ is expected to be positive if 
dependence in the left halves of the distributions is stronger 
than dependence in their right halves; likewise for $\Ps_l-\Ps_u$.
This holds, for instance, for the Clayton copula [@nelsen2007_copula].
The difference $\Dsq_u-\Dsq_l$, in which the "upper" and
"lower" statistics occur in opposite order, is also expected to be 
positive under the same conditions. We use these three differences
to detect asymmetry of tail dependence. Likewise, for large datasets,
we consider $\cor_{l_b,u_b}-\cor_{1-u_b,1-l_b}$, 
$\Ps_{l_b,u_b}-\Ps_{1-u_b,1-l_b}$ and
$\Dsq_{1-u_b,1-l_b}-\Dsq_{l_b,u_b}$ with $l_b=0$ and $u_b$ close to $0$.

In addition to [basic unit tests](https://github.com/sghosh89/BIVAN/blob/master/Tests.Rmd), we 
also tested our statistics by testing whether $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and 
$\Dsq_u-\Dsq_u$ revealed the asymmetry of tail dependence known to be present in data simulated
from various copulas. Normal and Frank copula families 
are known to have symmetric tail dependence,
i.e., strengths of dependence in the lower and upper tails are the same, for all copulas
in these families [@nelsen2007_copula]. Clayton copulas are known to have stronger lower- than 
upper-tail dependence. And survival Clayton copulas, being rotations of Clayton copulas by 
$180^\circ$, have stronger upper- than lower tail dependence [@nelsen2007_copula].
Using the `iRho` function from the `copula` package, for each of the four families normal, 
Frank, Clayton, and survival Clayton, we obtained parameters for which the expected value of 
Spearman's $\rho$ was, separately, $0,0.1,0.2,...,0.9$. For each copula family and for each 
of the selected parameter values we generated  $n$ points, 100 times, and thereby computed 100 values of the 
statistics $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_u$. This exercise 
was repeated, separately, 
for $n=1000$ and for $n=35$, these cases being chosen, respectively, to reflect asymptotic behavior of the 
statistics and their behavior on the small datasets which are common in some fields
of ecology. The mean and standard error of the three difference statistics were computed for 
each copula family, parameter value, and value of $n$, and results were plotted against 
Spearman's rho (Figures \ref{fig_stat_testing35} and \ref{fig_stat_testing1000}), 
the expectation being that the difference statistics would be positive for 
Clayton copulas and negative for survival Clayton copulas (except for the case of $\rho=0$, a 
boundary case of these families corresponding to independence), and not significantly different 
from $0$ (we used a $t$-test with a significance threshold of $0.05$) for Frank and normal 
copulas. These expectations were fullfilled (Figures \ref{fig_stat_testing35} and 
\ref{fig_stat_testing1000}).

@@

For multivariate datasets, e.g., the aphid abundance or first flight data, each of which consists of time series 
from 10 locations, a spatial resampling scheme was used instead of the surrogate approach described above.
The scheme is identical to that proposed by [@bjornstadfalck2001] for application in their now 
widely used spline correlogram method; the same scheme was also used by [@walteretal2017] as part of a method for 
performing model selection among matrix regression models. For each pair of locations $i \neq j$, the statistic
of interest, here denoted $s_{ij}$ ($\cor_{l_b,u_b}$, $\Ps_{l_b,u_b}$, $\Dsq_{l_b,u_b}$, $\cor_l-\cor_u$, 
$\Ps_l-\Ps_u$, or $\Dsq_u-\Dsq_l$) was computed for those locations. The values $s_{i,j}$ were then arranged in
a matrix, $S$, with diagonal entries NA. The mean of the non-NA entries of $S$ was computed. For 
$\cor_{l_b,u_b}$, $\Ps_{l_b,u_b}$, or $\Dsq_{l_b,u_b}$ this mean characterized the average strength of 
dependence, across all pairs of distinct locations, for data in the portion of the distributions 
given by $l_b$ and $u_b$. For $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, or $\Dsq_u-\Dsq_l$, the mean characterized average
asymmetry of tail dependence across all pairs of locations. 
Confidence intervals for the values of $\mean(S)$ were then computed as quantiles of a distribution of bootstrapped values.
Bootstrapping was performed by resampling the locations of measurement, with replacement; this corresponds to 
resampling, with replacement, the rows and columns of $S$, and is described in detail elsewhere 
[@bjornstadfalck2001 ; @walteretal2017].

@@

<!--Shya, can you please organize these following plots into one plot with mutliple panels? Once that is done, please add some text describing these results.-->
\begin{figure}[!h]
\begin{center}
\textbf{\;\;\;\;\;\;\;\ (A) \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\ (B)} \\
\includegraphics[width=2in]{./Results/stat_results/stat_aphid_count/Sp_10_Spearman_vs_D.pdf}
\includegraphics[width=2in]{./Results/stat_results/stat_aphid_count/Sp_10_Kendall_vs_D.pdf}\\
\textbf{\;\;\;\;\;\;\;\ (C) \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\ (D)} \\
\includegraphics[width=2in]{./Results/stat_results/stat_aphid_count/Sp_10_Corl-Coru_vs_D.pdf}
\includegraphics[width=2in]{./Results/stat_results/stat_aphid_count/Sp_10_Pl-Pu_vs_D.pdf}\\
\textbf{\;\;\;\;\;\;\;\ (E)} \\
\includegraphics[width=2in]{./Results/stat_results/stat_aphid_count/Sp_10_D2u-D2l_vs_D.pdf}
\caption{Results for Aphid-count}
\end{center}\label{fig_aphid_count}
\end{figure}

@@

<!--Shya, can you please organize these following plots into one plot with mutliple panels? Once that is done, please add some text describing these results.-->
\begin{figure}[!h]
\begin{center}
\textbf{\;\;\;\;\;\;\;\ (A) \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\ (B)} \\
\includegraphics[width=2in]{./Results/stat_results/stat_aphid_ff/Sp_11_Spearman_vs_D.pdf}
\includegraphics[width=2in]{./Results/stat_results/stat_aphid_ff/Sp_11_Kendall_vs_D.pdf}\\
\textbf{\;\;\;\;\;\;\;\ (C) \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\ (D)} \\
\includegraphics[width=2in]{./Results/stat_results/stat_aphid_ff/Sp_11_Corl-Coru_vs_D.pdf}
\includegraphics[width=2in]{./Results/stat_results/stat_aphid_ff/Sp_11_Pl-Pu_vs_D.pdf}\\
\textbf{\;\;\;\;\;\;\;\ (E)} \\
\includegraphics[width=2in]{./Results/stat_results/stat_aphid_ff/Sp_11_D2u-D2l_vs_D.pdf}
\caption{Results for Aphid-ff}
\end{center}\label{fig_aphid_ff}
\end{figure}

\newpage
## Results: abundance data for *Ceratium furca* in UK seas \label{Ex_pns}
